---
title: "Experimental prolong Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental prolong Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(prolong)
data("sim_outcome")
data("sim_metabs")
```


# Warning

These functions can produce additional visualizations but are not as soundly backed by our methodology and are subject to drastic change throughout versions of `prolong`.


## `delta_scatter()` Function

The `delta_scatter()` function can show a 2D or 3D scatterplot of pairwise delta-scale correlations to view the relationship between these pairwise correlations across time. The default is a 2D interactive scatterplot with hovertext for each point, and with fisher z-transformation applied to the pairwise correlations.

```{r, eval = F, warning=FALSE}
delta_scatter(sim_metabs)
```

If `interactive = FALSE` a non-interactive version will be shown without the hovertext. We can also change `timediff1` and `timediff2` to compare different time point pairs. If we set `fisherz = FALSE`, the raw pairwise correlations will be shown.

```{r, warning = F}
delta_scatter(sim_metabs, timediff1 = "3-2", timediff2 = '4-3', interactive = FALSE, fisherz = FALSE)
```

For 3D scatterplots, we can repeat the above but specify `timediff3`. Setting `interactive = FALSE` will not change anything if `timediff3` is supplied, only an interactive option is currently available.

```{r}
delta_scatter(sim_metabs, timediff3 = '4-3')
```




## `delta_network()` Function

The `delta_network()` function plots a network where the nodes are the $p$ variables and edges exist only between pairs of variables whose (partial) pearson correlation is greater than `corr_thresh`. The default is an interative network plot with edges between pairs of variables with partial correlation greater than $0.75$.

```{r,warning= F}
delta_network(sim_metabs)
```

As in `delta_scatter()` we can view a static plot instead with `interactive = FALSE`. We can also change the correlation threshold with `corr_thresh`, set `partial = FALSE` to use regular pairwise correlations, and change `method` to either `"kendall"` or `"spearman"` instead of the default `"pearson"`.

```{r}
delta_network(sim_metabs, timediff = "3-2", partial = FALSE, corr_thresh = 0.8, interactive = FALSE, method = "spearman")
```

