---
title: "Getting Started with prolong"
author: "Steve Broll"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with prolong}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r,eval = TRUE}
library(prolong)
```


The `prolong` package has two primary objectives:

* Facilitate the penalized modeling of high-dimensional longitudinal omics data with a longitudinal clinical outcome to obtain a sparse set of omics variables that co-vary with the outcome over time

* Provide functionality in both R and in a point-and-click GUI shiny app for plots that reveal the underlying dependence structure of the omics variables

For the first objective, the `prolong()` function takes an $n \times t$ matrix of clinical outcomes and a $n \times p \times t$ array of omics or omics-like variables and fits a model with a network penalty via laplacian matrix together with either a lasso or group lassa penalty. Hyperparameters can either be supplied or automatically estimated internally in `prolong()`. The `plot_trajectories()` function takes the output from `prolong()` and displays a grid of the trajectories of the variables selected by the (group) lasso + laplacian model. 

For the second objective, the `prolong` package will eventually sypport several plot functions that take the same $n \times p \times t$ array of omics or omics-like variables and look at the pairwise correlations of the first-differenced, or delta-scale, omics or omics-like data. The following function is currently supported: 

* `delta_heatmap()` - Interactive subsettable heatmaps

<!-- * `delta_scatter()` - 2D or 3D scatterplots  -->

<!-- * `delta_network()` - Interactive network visualizations -->

The `prolong` package comes with example data that is simulated to imitate the motivating data for the `prolong` methodology. There are two components that can be accessed with the `load()` function: 

* `sim_metabs` Simulated variables in a $15 \times 100 \times 4$ array

* `sim_outcome` Simulated outcome in a $15 \times 4$ matrix

# Example Data

The `prolong` package comes with example metabolite-inspired variables and viral load-inspired outcome simulated to imitate the real data that motivated `prolong`.

## Example Outcome

The example outcome data can be accessed with:


```{r, echo = T}
data("sim_outcome")
```

This simulated clinical outcome data is a $15 \times 4$ matrix, representing measurements of a continuous outcome for 15 subjects over 4 time points. 

```{r, echo=F}
dat <- matrix(NA, 60, 1)
for (i in 1:15) {
  dat[(i - 1) * 4 + seq(4), ] <- sim_outcome[i, ]
}
timelabs <- 1:4
time <- factor(rep(timelabs, 15), levels = timelabs)
id <- rep(1:15, each = 4)
colors <- "#00BFC4"
value <- reshape2::melt(dat)$value
meltdf <- data.frame(value, time, id, colors)
# head(meltdf)


ggplot2::ggplot(meltdf, ggplot2::aes(x = time, y = value, group = id)) +
  ggplot2::geom_line(color = colors) +
  ggplot2::theme(legend.position = "none") +
  ggplot2::xlab("Time") +
  ggplot2::ylab("Outcome")
```

## Example Variables

The example variable array can be accessed with:

```{r}
data("sim_metabs")
```


This simulated datasetis a $15 \times 100 \times 4$ matrix, representing measurements of 100 continuous variables for 15 subjects over 4 time points. This is the shape that `prolong` package functions expect the independent variabels to take.

The variables are labeled as target 1-20 and noise 1-80, with the targets correlated to each other and used to generate the outcome and the noise uncorrelated. The `plot_trajectories()` function can be used here to plot a sample of the variables' trajectories.

```{r}
plot_trajectories(sim_metabs, selected = 1:20)
```

# `prolong()` Function

Now that we have loaded our x array and y matrix, we can fit a lasso + laplacian and a group lasso + laplacian model. To fit the (non-group) lasso + laplacian model, we can specify `groups = FALSE`. If we don't specify any of the model hyperparameters they will be automatically found.

```{r, cache = F}
promod_ll <- prolong(sim_metabs, sim_outcome, groups = FALSE)
```

Now that we have the results, we can quickly and easily visualize the selected variables with `plot_trajectories()`.

```{r}
plot_trajectories(sim_metabs, promod_ll)
```

Let's repeat the process but with the group lasso + laplacian model. The default is to automatically create 1 group for each of the $p$ variables, with the different time point observations as replications within that group.

```{r, cache = F}
promod_gll <- prolong(sim_metabs, sim_outcome)
```


```{r}
plot_trajectories(sim_metabs, promod_gll)
```

# Correlation Plots


## `delta_heatmap()` Function

The `delta_heatmap()` function takes the $n\times p \times t$ array of variables and produces a heatmap of the delta-scale absolute correlations to visualize  By default, `delta_heatmap()` will open a window to display an interactive Shiny app with the option to view a selected subsection of the heatmap.

```{r, eval = FALSE, message = FALSE, warning=FALSE}
delta_heatmap(sim_metabs)
```

If `interactive = FALSE`, then a static heatmap will instead be produced. We can also change the time points whose difference we take before computing the absolute correlation matrix.


```{r}
delta_heatmap(sim_metabs, timediff = "3-2", interactive = F)
```
