---
title: "prolong"
author: "Steve Broll"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{prolong}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r,eval = FALSE}
library(prolong)

```


The `prolong` package has two primary objectives:

* Facilitate the penalized modeling of high-dimensional longitudinal omics data with a longitudinal clinical outcome to obtain a sparse set of omics variables that co-vary with the outcome over time

* Provide functionality in both R and in a point-and-click GUI shiny app for plots that reveal the underlying dependence structure of the omics variables

For the first objective, the `prolong()` function takes an $n \times t$ matrix of clinical outcomes and a $n \times p \times t$ array of omics or omics-like variables and fits a model with a network penalty via laplacian matrix together with either a lasso or group lassa penalty. Hyperparameters can either be supplied or automatically estimated internally in `prolong()`. The `plot_trajectories()` function takes the output from `prolong()` and displays a grid of the trajectories of the variables selected by the (group) lasso + laplacian model. 

For the second objective, the `prolong` package has several plot functions that take the same $n \times p \times t$ array of omics or omics-like variables and look at the pairwise correlations of the first-differenced, or delta-scale, omics or omics-like data. The following functions are currently supported: 

* `delta_scatter()` - 2D or 3D scatterplots 

* `delta_heatmap()` - Interactive subsettable heatmaps

* `delta_network()` - Interactive network visualizations

The `prolong` package comes with example data that is simulated to imitate the motivating data for the `prolong` methodology. There are two components that can be accessed with the `load()` function: 

* `sim_metabs` Simulated variables in a $15 \times 100 \times 4$ array

* `sim_outcome` Simulated outcome in a $15 \times 4$ matrix

